---
title: "Ellsworth Resurvey Data Analysis"
output: github_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

## Goals and Approach

The primary goal of these preliminary analyses is to figure out how many plots should be resurveyed at Ellsowrth in summe, echo=TRUE, results="hide"r 2020. To do this, we used the pre-treatment survey data from 2006/2007 to quantify variance in density, height, and DBH of plots by  block and stand type. 

## Summary 

## Data Files
We used the following datafiles, which can be in the Ellsworth GitHub repo (https://github.com/AileneKane/ellsworth) as well as in the "Ellsworth Science and Monitoring" shared folder on Box.

LIVETREES_CLEAN_04282008.csv

PLOT_CLEAN_11062008.csv

Ellsworth_stands_treatment_data.csv

## Analysis of pre-treatment survey data

We load the above files and packages, and then put them together into a dataframe for analysis
```{r getdata, include = FALSE}
rm(list=ls()) 
options(stringsAsFactors = FALSE)

# libraries
library(dplyr)
library(lme4)
library(scales)
#set working directory
setwd("~/GitHub/ellsworth")

#1. read in tree dat, plot data, and treatment data
treed<-read.csv("data/LIVETREES_CLEAN_04282008.csv", header=TRUE)
plotd<-read.csv("data/PLOT_CLEAN_11062008.csv", header=TRUE)
trtd<-read.csv("data/Ellsworth_stands_treatment_data.csv", header=TRUE)
#2. put the data together and standardize some columns for model fitting
source("analyses/source/prep_data.R")


```
There is some structure to the data, that may be important to account for in looking at treatment effects, and therefore should be kept in mind in resurveys: 
-Blocks (N,C,S), which are different geographical regions at Ellsworth
-Standtype, which are different forest types at Ellsworth.
Here is a breakdown of the number of plots in each standtype by region:

```{r showdata, include = TRUE, echo = TRUE}
 table(plotd2$STAND.TYP,plotd2$BLOCK)

```

Now lets look at effects of age on density, height, dbh, and crown, and quantify variation by region and stand-type using multilevel models.

## Density

```{r densmod, include = TRUE, echo = TRUE}
plotd2$stand.code<-substr(plotd2$STAND.TYPE,1,2)
colors<-c("darkblue","lightgreen","goldenrod")
symbs<-c(21,22,23,24)
blocks<-as.character(unique(plotd2$BLOCK))
treats<-unique(plotd2$TRT)
sttypes<-sort(unique(plotd2$stand.code))

plot(plotd2$AGE_BH_2006,plotd2$predens, pch=symbs[as.numeric(as.factor(plotd2$stand.code))], bg=colors[as.numeric(as.factor(plotd2$BLOCK))], xlab="Age, 2006 (years)", ylab= "Density, 2006 (trees/plot)", main="density vs. age, by block and stand-type")

legend("topright", legend=c(blocks,sttypes),pch=c(22,22,22,symbs),pt.bg=c(colors,"lightgreen","lightgreen","lightgreen","lightgreen") )

densmod<-lmer(predens~AGE_BH_2006 + (1|BLOCK)  + (1|STAND.TYPE), data=plotd2)
den.standsd<-VarCorr(densmod,comp="Variance")[1]
den.blocksd<-VarCorr(densmod,comp="Variance")[2]
summary(densmod)#
```
There is a weak negative effect of age on density,  after accounting for variation in density and blocks. There is much higher variance in density by stand type than by block.

## DBH
```{r dbh, include = TRUE, echo = TRUE}

plot(plotd2$AGE_BH_2006,plotd2$dbh.mn, pch=symbs[as.numeric(as.factor(plotd2$stand.code))], bg=colors[as.numeric(as.factor(plotd2$BLOCK))], xlab="Age, 2006 (years)", ylab= "Mean DBH, 2006", main="DBH vs. age, by block and stand-type")

legend("bottomright", legend=c(blocks,sttypes),pch=c(22,22,22,symbs),pt.bg=c(colors,"lightgreen","lightgreen","lightgreen","lightgreen") )
dbhmod<-lmer(DBH~ AGE_BH_2006+(1|BLOCK) + (1|STAND.TYPE), data=treed2)
dbh.standsd<-VarCorr(dbhmod,comp="Variance")[1]
dbh.blocksd<-VarCorr(dbhmod,comp="Variance")[2]
summary(dbhmod)#positive effect of age on dbh, similar variance by TRT and Block
```
There is a positive effect of age on dbh,  after accounting for variation in density and blocks. There is much higher variance in dbh by stand type than by block.

## Height
```{r height, include = TRUE, echo = TRUE}


plot(plotd2$AGE_BH_2006,plotd2$ht.mn, pch=symbs[as.numeric(as.factor(plotd2$stand.code))], bg=colors[as.numeric(as.factor(plotd2$BLOCK))], xlab="Age, 2006 (years)", ylab= "MEAN HEIGHT, 2006", main="HEIGHT vs. age, by block and stand-type")

legend("bottomright", legend=c(blocks,sttypes),pch=c(22,22,22,symbs),pt.bg=c(colors,"lightgreen","lightgreen","lightgreen","lightgreen") )
htmod<-lmer(HT~ AGE_BH_2006+(1|BLOCK) + (1|STAND.TYPE), data=treed2)
summary(htmod)#positive effect of age on ht, higher variance by TRT...
ht.standsd<-VarCorr(htmod,comp="Variance")[1]
ht.blocksd<-VarCorr(htmod,comp="Variance")[2]
```
There is a positive effect of age on height,  after accounting for variation in density and blocks. There is much higher variance in height by stand type than by block.
## Crown

```{r crownmod, include = TRUE, echo = TRUE}
plot(plotd2$AGE_BH_2006,plotd2$crown.mn, pch=symbs[as.numeric(as.factor(plotd2$stand.code))], bg=colors[as.numeric(as.factor(plotd2$BLOCK))], xlab="Age, 2006 (years)", ylab= "MEAN CROWN, 2006", main="CROWN vs. age, by block and stand-type")

legend("bottomright", legend=c(blocks,sttypes),pch=c(22,22,22,symbs),pt.bg=c(colors,"lightgreen","lightgreen","lightgreen","lightgreen"))

crownmod<-lmer(CROWN~ AGE_BH_2006+(1|BLOCK) + (1|STAND.TYPE), data=treed2)

summary(crownmod)#positive effect of age on crown, lsightly higher variance in TRt than block
crown.standsd<-VarCorr(crownmod,comp="Variance")[1]
crown.blocksd<-VarCorr(crownmod,comp="Variance")[2]

```
There is a positive effect of age on crown,  after accounting for variation in density and blocks. There is much higher variance in crown by stand type than by block.


## How many plots do we need to capture treatment effects, amidst all the variation?
