---
title: "Ellsworth Resurvey Data Analysis"
output: github_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

## Goals and Approach

The primary goal of these preliminary analyses is to figure out how many plots should be resurveyed at Ellsowrth in summe, echo=TRUE, results="hide"r 2020. To do this, we used the pre-treatment survey data from 2006/2007 to quantify variance in density, height, and DBH of plots by  block and stand type. 

## Summary 

## Data Files
We used the following datafiles, which can be in the Ellsworth GitHub repo (https://github.com/AileneKane/ellsworth) as well as in the "Ellsworth Science and Monitoring" shared folder on Box.

LIVETREES_CLEAN_04282008.csv

PLOT_CLEAN_11062008.csv

Ellsworth_stands_treatment_data.csv

## Analysis of pre-treatment survey data

We load the above files and packages, and then put them together into a dataframe for analysis
```{r getdata, include = FALSE}
rm(list=ls()) 
options(stringsAsFactors = FALSE)

# libraries
library(dplyr)
library(lme4)
library(scales)
#set working directory
setwd("~/GitHub/ellsworth")

#1. read in tree dat, plot data, and treatment data
treed<-read.csv("data/LIVETREES_CLEAN_04282008.csv", header=TRUE)
plotd<-read.csv("data/PLOT_CLEAN_11062008.csv", header=TRUE)
trtd<-read.csv("data/Ellsworth_stands_treatment_data.csv", header=TRUE)
#2. put the data together and standardize some columns for model fitting
source("analyses/source/prep_data.R")


```
There is some structure to the data, that may be important to account for in looking at treatment effects, and therefore should be kept in mind in resurveys: 
-Blocks (N,C,S), which are different geographical regions at Ellsworth
-Standtype, which are different forest types at Ellsworth.
Here is a breakdown of the number of plots in each standtype by region:

```{r showdata, include = TRUE, echo = TRUE}
 table(plotd2$STAND.TYP,plotd2$BLOCK)

```

Now lets look at effects of age on density, height, dbh, and crown, and quantify variation by region and stand-type using multilevel models.

## Density

```{r densmod, include = TRUE, echo = TRUE}
colors<-c("darkblue","lightgreen","goldenrod")
symbs<-c(21,22,23,24)
blocks<-as.character(unique(plotd2$BLOCK))
treats<-unique(plotd2$TRT)
sttypes<-sort(unique(plotd2$stand.code))

plot(plotd2$AGE_BH_2006,plotd2$predens, pch=symbs[as.numeric(as.factor(plotd2$stand.code))], bg=colors[as.numeric(as.factor(plotd2$BLOCK))], xlab="Age, 2006 (years)", ylab= "Density, 2006 (trees/plot)", main="density vs. age, by block and stand-type")

legend("topright", legend=c(blocks,sttypes),pch=c(22,22,22,symbs),pt.bg=c(colors,"lightgreen","lightgreen","lightgreen","lightgreen") )

densmod<-lmer(predens~AGE_BH_2006 + (1|BLOCK)  + (1|STAND.TYPE), data=plotd2)

den.standsd<-VarCorr(densmod,comp="Variance")[1]
den.blocksd<-VarCorr(densmod,comp="Variance")[2]
Bage.dens<-fixef(densmod)[2]
int.dens<-fixef(densmod)[1]
summary(densmod)#
```
There is a weak negative effect of age on density,  after accounting for variation in density and blocks. There is much higher variance in density by stand type than by block.

## DBH
```{r dbh, include = TRUE, echo = TRUE}

plot(plotd2$AGE_BH_2006,plotd2$dbh.mn, pch=symbs[as.numeric(as.factor(plotd2$stand.code))], bg=colors[as.numeric(as.factor(plotd2$BLOCK))], xlab="Age, 2006 (years)", ylab= "Mean DBH, 2006", main="DBH vs. age, by block and stand-type")

legend("bottomright", legend=c(blocks,sttypes),pch=c(22,22,22,symbs),pt.bg=c(colors,"lightgreen","lightgreen","lightgreen","lightgreen") )
dbhmod<-lmer(DBH~ AGE_BH_2006+(1|BLOCK) + (1|STAND.TYPE), data=treed2)
dbh.standsd<-VarCorr(dbhmod,comp="Variance")[1]
dbh.blocksd<-VarCorr(dbhmod,comp="Variance")[2]
summary(dbhmod)#positive effect of age on dbh, similar variance by TRT and Block
```
There is a positive effect of age on dbh,  after accounting for variation in density and blocks. There is much higher variance in dbh by stand type than by block.

## Height
```{r height, include = TRUE, echo = TRUE}


plot(plotd2$AGE_BH_2006,plotd2$ht.mn, pch=symbs[as.numeric(as.factor(plotd2$stand.code))], bg=colors[as.numeric(as.factor(plotd2$BLOCK))], xlab="Age, 2006 (years)", ylab= "MEAN HEIGHT, 2006", main="HEIGHT vs. age, by block and stand-type")

legend("bottomright", legend=c(blocks,sttypes),pch=c(22,22,22,symbs),pt.bg=c(colors,"lightgreen","lightgreen","lightgreen","lightgreen") )
htmod<-lmer(HT~ AGE_BH_2006+(1|BLOCK) + (1|STAND.TYPE), data=treed2)
summary(htmod)#positive effect of age on ht, higher variance by TRT...
ht.standsd<-VarCorr(htmod,comp="Variance")[1]
ht.blocksd<-VarCorr(htmod,comp="Variance")[2]
```
There is a positive effect of age on height,  after accounting for variation in density and blocks. There is much higher variance in height by stand type than by block.
## Crown

```{r crownmod, include = TRUE, echo = TRUE}
plot(plotd2$AGE_BH_2006,plotd2$crown.mn, pch=symbs[as.numeric(as.factor(plotd2$stand.code))], bg=colors[as.numeric(as.factor(plotd2$BLOCK))], xlab="Age, 2006 (years)", ylab= "MEAN CROWN, 2006", main="CROWN vs. age, by block and stand-type")

legend("bottomright", legend=c(blocks,sttypes),pch=c(22,22,22,symbs),pt.bg=c(colors,"lightgreen","lightgreen","lightgreen","lightgreen"))

crownmod<-lmer(CROWN~ AGE_BH_2006+(1|BLOCK) + (1|STAND.TYPE), data=treed2)

summary(crownmod)#positive effect of age on crown, lsightly higher variance in TRt than block
crown.standsd<-VarCorr(crownmod,comp="Variance")[1]
crown.blocksd<-VarCorr(crownmod,comp="Variance")[2]

```
There is a positive effect of age on crown,  after accounting for variation in density and blocks. There is much higher variance in crown by stand type than by block.


## How many plots do we need to capture treatment effects, amidst all the variation?

To figure out how many plots need to  be resampled to capture the treatment effects, amidst all the variation among blocks and standtypes, we simulated a dataset of resampled data. To do this, we set expected effect sizes of treatment
```{r simdat, include = TRUE, echo = TRUE}
#4. set expected effect sizes for simulating data
dens.b = 1#effect of pre-density 
age.b = Bage.dens#age effect (older stands have lower post-density? use coef from density model above)
trt.b = -1 #trt is amount removed (0,3 x predens)
sigma.bl =6.055#estimated block-level sigma from 2006 data 
sigma.st =22.37#estimated stand-type sigma from 2006 data 

sigma = .1#
b0 = int.dens# from model above

#use the effect sizes and predictors plus error to generate the y variable

block = x$block
nblock = length(unique(x$block))
nplot = 72

nstand = 3
blockeff = rep(rnorm(nblock, 0, sigma.bl), each = nplot)
standeff = rep(rnorm(nstand, 0, sigma.st), each = nplot)
ploteff = rnorm(nblock*nplot, 0, sigma)

ypred = b0 + dens.b*x$predens.z + age.b*x$age2006.z + trt.b*x$trt.z+ blockeff+ standeff + ploteff

#lmer(ypred ~ predens.z + age2006.z + trt.z+ (1|block)+ (1|stand.code), data=x )


#now lets write a for loop that uses different sample sizes to figure out how many plots are needed to correctly recover the effects
fulldat<-cbind(ypred,x)
nplots<-rep(c(5,10,15,20,25,30,35,40,45), times=10)
allplots<-c()
for(i in 1:length(nplots)){
  subsdatc<-sample_n(fulldat[fulldat$block=="C",], nplots[i])
  subsdatn<-sample_n(fulldat[fulldat$block=="N",], nplots[i])
  subsdats<-sample_n(fulldat[fulldat$block=="S",], nplots[i])
  subsdat<-rbind(subsdatc,subsdatn,subsdats)
  fit<-lmer(ypred ~ predens.z + age2006.z + trt.z+ (1|block)+ (1|stand.code),data=subsdat)
  cis<-confint(fit)
  fit.sum<-c(nplots[i],fixef(fit),cis[3,],cis[4,],cis[5,],cis[6,])
  allplots<-rbind(allplots,fit.sum)
}
allplots<-as.data.frame(allplots)
colnames(allplots)<-c("n","int","predens.b","age.b","trt.b","int.lc","int.uc","predens.b.lc","predens.b.uc","age.b.lc","age.b.uc","trt.b.lc","trt.b.uc")


par(mfrow=c(1,3))
plot(allplots$n,allplots$predens.b,main="pre-density",ylim=c(0,7))
for(i in 1:dim(allplots)[1]){
  arrows(allplots$n[i],allplots$predens.b.lc[i],allplots$n[i],allplots$predens.b.uc[i],length=0.1,code = 0, angle= 90, lwd=3,col= alpha("gray",0.1))
}
abline(h=dens.b, lwd=2, col="red")
plot(allplots$n,allplots$age.b,main="age",ylim=c(-1,0))
for(i in 1:dim(allplots)[1]){
  arrows(allplots$n[i],allplots$age.b.lc[i],allplots$n[i],allplots$age.b.uc[i],length=0.1,code = 0, angle = 90, lwd=3,col= alpha("gray",0.1))
}
abline(h=age.b, lwd=2, col="red")
plot(allplots$n,allplots$trt.b,main="trt",ylim=c(-1.5,0))
for(i in 1:dim(allplots)[1]){
  arrows(allplots$n[i],allplots$trt.b.lc[i],allplots$n[i],allplots$trt.b.uc[i],length=0.1,code = 0, angle = 90, lwd=3,col= alpha("gray",0.1))
}
abline(h=trt.b, lwd=2, col="red")
```


